BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-06-25 LMThesisNicchi.tex
Thread Management Section Completed

BlueTracer has been organized primarily taking into account the rich set of APIs offered by Pin, which have led to the decision to split the tool in three parts: the first aimed at \textbf{native API tracing}, the second for \textbf{API tracing} and the last focused on \textbf{callbacks and APCs tracing} (\textit{Figure 3.1}). In particular, the tracing of native APIs also employs a different source of API information (\textit{Dr. Memory}'s system call data) than the ones utilized for tracing APIs (\textit{drltrace}'s configuration file or the information extracted from \textit{Pyrebox}'s database).

@@

Pin assigns to each thread an ID, a small number beginning at 0
which is not the same as the operating system thread ID. A way to obtain such ID is by using as analysis routine argument \texttt{IARG\_THREAD\_ID}, which passes the thread ID assigned by Pin for the calling thread. This ID is typically used as an index of an array of thread data.
In fact, the Pin API makes available an efficient thread local storage (TLS). In order to utilize it, it is first required to allocate a new TLS key via \texttt{PIN\_CreateThreadDataKey}, which can optionally take as input a pointer to a destructor function. After that, any thread of the process can use the TLS key, in addition to its Pin-specific thread ID, to store (\texttt{PIN\_SetThreadData}) and retrieve (\texttt{PIN\_GetThreadData}) values in its own slot. The starting value relative to the key in every thread is \texttt{NULL}.
Pin makes also available call-backs when each thread starts (\texttt{PIN\_AddThreadStartFunction}) and ends (\texttt{PIN\_AddThreadFiniFunction}). This is typically where thread local data is allocated, manipulated and stored in a thread's local storage\cite{Pin}.

In BlueTracer the TLS for each thread is as follows. Explain most important fields. Allocated in Onthreadstart and deallocated in thread exit.

\vspace{0.5cm}
\begin{lstlisting}[caption={My Caption},captionpos=b]
typedef struct {
 ...
 syscall_tracer* syscallEntry;     // Pointer to NTAPI entry
 vector<stackEntry>* shadowStack;  // Shadow stack
 uint call_number;             	   // Calls counter

@@

\subsection{File Management}

\section{NTAPI}
\iffalse
Talk about log structure when talking about how arguments are printed
\fi

>>>>

BlueTracer has been organized primarily taking into account the rich set of APIs offered by Pin, which have led to the decision to split the tool in three parts: the first aimed at \textbf{native APIs tracing}, the second for \textbf{APIs tracing} and the last focused on \textbf{callbacks and APCs tracing} (\textit{Figure 3.1}). In particular, the tracing of native APIs also employs a different source of API information (\textit{Dr. Memory}'s system call data) than the ones utilized for tracing APIs (\textit{drltrace}'s configuration file or the information extracted from \textit{Pyrebox}'s database).

@@

Pin assigns to each thread an ID, a small number beginning at 0
which is not the same as the operating system thread ID. A way to obtain such ID is by using as analysis routine argument \texttt{IARG\_THREAD\_ID}, which passes the thread ID assigned by Pin for the calling thread. This ID is typically used as an index of an array of thread data.
In fact, the Pin API makes available an efficient thread local storage (TLS). In order to utilize it, it is first required to allocate a new TLS key via \texttt{PIN\_CreateThreadDataKey}, which can optionally take as input a pointer to a destructor function. After that, any thread of the process can use the TLS key, in addition to its Pin-specific thread ID, to store (\texttt{PIN\_SetThreadData}) and retrieve (\texttt{PIN\_GetThreadData}) values in its own slot. The starting value relative to the key in every thread is \texttt{NULL}.
Pin makes also available call-backs when each thread starts (registered with \texttt{PIN\_AddThreadStartFunction}) and ends (registered with \texttt{PIN\_AddThreadFiniFunction}). This is typically where thread local data is allocated, manipulated and stored in a thread's local storage\cite{Pin}.

In BlueTracer, each TLS slot stores a \texttt{struct} of the type \texttt{bluepill\_tls} (\textit{Listing 3.1}) for every thread. Such \texttt{struct} is dynamically allocated every time a thread starts in the \texttt{OnThreadStart} callback function and is consequently deallocated in the \texttt{onThreadFini} function when the thread ends.

\vspace{0.5cm}
\begin{lstlisting}[caption={Thread Local Data},captionpos=b]
typedef struct {
 ...
 syscall_tracer* syscallEntry;     // Pointer to NTAPI entry
 vector<stackEntry>* shadowStack;  // Shadow stack
 uint call_number;             	   // Calls counter

@@

Since the first three fields of the above \texttt{struct} (lines 3-5) are employed when tracing native APIs and APIs, they will be discussed in detail later in the chapter. Now let us focus on the remaining fields, which are used by BlueTracer to write the traced information in the appropriate log files.

\subsection{Log Files and Multithreading}

In BlueTracer, the traced data is written to a binary file, one for every thread. The default naming convention used for the tracer's log files is \texttt{Traced.[OS Process ID].[Pin-specific Thread ID]}, similarly to the one Blue Pill employs in its own log files, with the user being able to change \texttt{Traced} with a name of its choice in the configuration file.

When writing data to file, each thread invokes the \texttt{file\_write} function, whose pointer is located in the instance of the \texttt{bluepill\_tls} struct associated to the thread (line 11 of \textit{Listing 3.1}). However, such data, which follows the same format of strings used by \texttt{fprintf}, is not directly written to file. Instead, an intermediate 8 kB buffer is used (line 7 of \textit{Listing 3.1}): only when the buffer is full (or when the amount of data to be written does not fit the buffer) file writing actually occurs. The choice of using a buffer was made as an attempt to improve performance, as it allows the aggregation of small write operations into a block size that is more efficient for the disk subsystem.

A problem which was encountered when trying to conjugate file management and multithreading is that there exists a known isolation issue affecting Pin on Windows. Specifically, it is possible for a deadlock to take place if a file is opened in a callback in the context of multithreaded applications. As a result of this issue, it is not possible to open the tracer's log file in the \texttt{OnThreadStart} callback. Pin's manual proposes to circumvent the problem by opening the file in the \texttt{main} and tagging the data with the thread ID \cite{Pin}. However, this conflicts with the idea of having one file for each thread.

In order to bypass this limitation of the Pin's framework, the following strategy was employed:
\begin{enumerate}
\item When initializing the thread local data in \texttt{OnThreadStart}, \texttt{file\_write} is set to point to a function named \texttt{file\_open}.
\item The first time a thread attempts to write data to file \texttt{file\_open} is invoked.
\\
\\
\item \texttt{file\_open} carries out the following actions:
\begin{enumerate}
\item Opens the tracer's log file (this is safe since the file is not opened in a callback)
\item Sets the obtained file pointer in the thread local data (line 8 of \textit{Listing 3.1})
\item Adds the data to be written in the buffer (which is eventually written to file if the buffer is too small to hold it)
\item Sets \texttt{file\_write} to point to \texttt{buf\_write}, a function which is in charge of just writing data to the buffer and to file.
\end{enumerate}
\item As a result of this, when the thread attempts to write to file again, \texttt{buf\_write} is invoked, thus allowing the thread to just write to file without going through opening the file again.   
\end{enumerate}

\section{Native APIs Tracing}
\iffalse
Talk about log structure when talking about how arguments are printed
\fi
